Gap formation and stability in non-isothermal protoplanetary discs

### 2014-08-07 paper/model.tex
paper update

Note that the energy source terms have been chosen to be absent 
at $t=0$. This allows the steady state energy equation to be satisfied
provided the flow is axisymmetric and $v_r$ is negligible compared to $v_\phi$.  
% If the disc is axisymmetric, steady and with only azimuthal flow, then the 
%energy equation is trivially satisfied. 
%This allows one to specify equilibrium discs (see below)
%in the same way as in the case where the energy equation is replaced by a fixed temperature 
%profile (i.e. $c_\mathrm{iso}^2\equiv p/\Sigma$ is prescribed). 
The cooling term attempts 
to restore the initial energy density (and therefore temperature) profile.

@@

\section{Numerical experiments}\label{method}
The self-gravitating disc-planet system is evolved using the 
\texttt{FARGO-ADSG} code \citep{baruteau08, baruteau08b}. This is a modified version 
of the original \texttt{FARGO} code \citep{masset00a} to include the energy 
equation. The code employs finite-difference scheme similar 
to the \texttt{ZEUS} code \citep{stone92}, but with a modified azimuthal transport 
algorithm to circumvent the time-step restriction set by the inner disc boundary. 
The disc is divided into $(N_r,N_\phi)$ zones in the radial and azimuthal directions, 
respectively. The grid spacing is logarithmic in radius and uniform in azimuth.

>>>>

Note that the energy source terms have been chosen to be absent 
at $t=0$. This allows the disc to be initialised in a steady state
energy, provided the flow is axisymmetric and $v_r$ is negligible compared to $v_\phi$.  
% If the disc is axisymmetric, steady and with only azimuthal flow, then the 
%energy equation is trivially satisfied. 
%This allows one to specify equilibrium discs (see below)
%in the same way as in the case where the energy equation is replaced by a fixed temperature 
%profile (i.e. $c_\mathrm{iso}^2\equiv p/\Sigma$ is prescribed). 
The energy source term attempts to restore the initial energy density (and
therefore temperature) profile. In practice, this is a cooling term at
the gap edge because disc-planet interaction leads to heating.

@@

\section{Numerical experiments}\label{method}
The disc-planet system is evolved using the 
\texttt{FARGO-ADSG} code \citep{baruteau08, baruteau08b}. This is a modified version 
of the original \texttt{FARGO} code \citep{masset00a} to include the energy 
equation. The code employs finite-difference scheme similar 
to the \texttt{ZEUS} code \citep{stone92}, but with a modified azimuthal transport 
algorithm to circumvent the time-step restriction set by the inner disc boundary. 
The disc is divided into $(N_r,N_\phi)$ zones in the radial and azimuthal directions, 
respectively. The grid spacing is logarithmic in radius and uniform in azimuth.

### 2014-08-07 paper/results.tex
paper update

\section{Numerical linear stability simulations}
We first consider the effect of a cooling time on the \emph{linear} stability
of planetary gaps. There are two competing effects. Increasing
$t_\mathrm{cool}$ leads to more heating and hence a larger sound-speed
$c_s$, which favours instability \citep{li00} since the vortex
instability is pressre-driven. However, gap-opening requires $r_h>H$,
so increasing the sound-speed, and hence $H$, leads to a shallower
gap (see below) and expected to be more stable.

In the following simulations, the planet is introduced at $t=20P_0$ and
its potential switched on over $10P_0$. At $t=30P_0$ we switch off the
planet potential and azimuthally average the surface density, energy
and velocity fields. We then perturb the density field and continue
evolve the disc. %more details on perturbation: m=1 to 10, amplitude?  
%resolution?
%switch off at t=30 because in full disk-planet sims, instability sets
%in soon after planet introduced

@@

\subsection{Linear modes and growth rates}
{\bf present `planet-off' simulations. one `fourier mode v.s. time'
  plot to show growth of linear instability. 
  compare growth rate and dominant
  m as function of beta (table). 2D figs to contrast (also used to
  show it's the minimum in generalized pv that goes unstable.    
  result: increasing cooling time makes the gap
  more stable, and favors lower m. note: the `basic state' should be
  the system at t=30 after azimuthal average. linear results should
  have small perturbations. 
}

%% We first describe results for weakly self-gravitating
%% discs $Q_0=8$ (giving $M_d=0.015M_*$) so that the gap remains 
%% gravitationally stable \citep{lin11b}. We also impose a dimensionless
%% kinematic viscosity $\hat{\nu}=2\times10^{-5}$ to suppress the Rossby
%% vortex instablity \citep{valborro07}. These parameter values
%% are typical for disc-planet simulations and yield stable/steady 
%% gap profiles. This will be useful reference cases to  
%% understand the unstable cases considered later.

>>>>

\section{Growth of non-axisymmetric modes without the influence of the
  planet}
%We first consider the linear stability of planetary gaps.  
In this section, the planet is introduced at $t=20P_0$ and 
its potential switched on over $10P_0$. At $t=30P_0$ we switch off the
planet potential and azimuthally average the surface density, energy
and velocity fields. At this point the planet has carved a partial
gap. We then perturb the density field and continue to 
evolve the disc. This procedure allows us to analyse the growth of
non-axisymmetric modes associated with the gap, but without
complications from non-axisymmetry arising directly from disk-planet
interaction.

%The goal of these simulations is to compliment the disk-planet
%simluations presented later. 
These `planet-off' simulations are not linear stability calculations
because our cooling term restores the initial temperature profile
corresponding to $H/r=0.05$, rather than the heated gap edge.

%disk-planet leads to heating 
%tcool tries to return it to the initial smooth profile (axisymmetric)
%if tcool > trwi -> temp returns to original value. heated gap not
%captured

%There are two competing effects. Increasing
%$t_\mathrm{cool}$ leads to more heating and hence a larger sound-speed
%$c_s$, which favours instability \citep{li00} since the vortex
%instability is pressre-driven. However, gap-opening requires $r_h>H$,
%so increasing the sound-speed, and hence $H$, leads to a shallower
%gap (see below) and expected to be more stable.

%tcool -> return to initial, t=0 temp. profile
%whereas the `basic state' is the gap profile with structured temp profile

%more details on perturbation: m=1 to 10, amplitude?  
%resolution?

@@

\subsection{Linear modes and growth rates}
{\bf present `planet-off' simulations. one `fourier mode v.s. time'
  plot to show growth of linear instability. do an adiabatic case (or
  extremely long cooling time) to see the effect of heated gap edge
  (i.e. temp doesn't go back to t=0 value too quickly)
  compare growth rate and dominant
  m as function of beta (table). 2D figs to contrast (also used to
  show it's the minimum in generalized pv that goes unstable.    
  result: increasing cooling time makes the gap
  more stable, and favors lower m. note: the `basic state' should be
  the system at t=30 after azimuthal average. linear results should
  have small perturbations. 
}

\subsubsection{Nearly adiabatic disks}
{\bf a case with extrmely long cooling time (or adiabatic run) in
  order to look at effect of heated gap edge.  prelim result: not
  important, at t=30, gap edge only heats to $H/r\simeq0.06$ even for
  purely adiabatic disk. change in $h$ not important for linear
  perturbations (but important for setting up the basic state). 
}
%% We first describe results for weakly self-gravitating
%% discs $Q_0=8$ (giving $M_d=0.015M_*$) so that the gap remains 
%% gravitationally stable \citep{lin11b}. We also impose a dimensionless
%% kinematic viscosity $\hat{\nu}=2\times10^{-5}$ to suppress the Rossby
%% vortex instablity \citep{valborro07}. These parameter values
%% are typical for disc-planet simulations and yield stable/steady 
%% gap profiles. This will be useful reference cases to  
%% understand the unstable cases considered later.
